---
title: "Reproducible Research - Course Project I"
output: html_document
---

## Loading & Preprocessing of Data

```{r setup}
#removes all variables from environment
rm(list=ls(all=TRUE)) 
```

Proceeding with downloading of data for this assignment 

Variables of downloaded dataset-
1. steps: Number of steps in 5-minute interval 
2. date: Date of measurement (YYYY-MM-DD format)
3. interval: Measurement identifier for 5-minute interval


```{r}
dat = read.csv('activity.csv', header = T)
names(dat)
str(dat)
head(dat)
```


## Histogram, Mean & Median

```{r}
library(data.table)
dat_tbl = data.table(dat)
dat_tbl_summary = dat_tbl[, list(total_steps = sum(steps, na.rm = T)), 
                          by = date]

```{r}

gen_hist = function(x, title){
        hist(x, 
             breaks = 20,
             main = title,
             xlab = 'Total Number of Steps', col = 'grey',
             cex.main = .9 
        mean_value = round(mean(x), 1)
        median_value = round(median(x), 1)
        abline(v=mean_value, lwd = 3, col = 'blue')
        abline(v=median_value, lwd = 3, col = 'red')
        legend('topright', lty = 1, lwd = 3, col = c("blue", "red"),
               cex = .8, 
               legend = c(paste('Mean: ', mean_value),
               paste('Median: ', median_value))
               )
}

gen_hist(dat_tbl_summary$total_steps, 'Number of Steps Taken Per Day')
```




## Daily Activity Pattern

----------------------------

The below plot addresses the following items:

Time series plot of 5-minute interval vs average number of steps 
Maximum number of steps


```{r}
dat_tbl_summary_intv = dat_tbl[, list(avg_steps = mean(steps, na.rm = T)), 
                          by = interval]
with(dat_tbl_summary_intv, {
        plot(interval, avg_steps, type = 'l',
             main = 'Average Steps by Time Interval',
             xlab = '5 Minute Time Interval',
             ylab = 'Average Number of Steps')
        })
#Find Interval That Has The Maximum Avg Steps
max_steps = dat_tbl_summary_intv[which.max(avg_steps), ]

max_lab = paste('Maximum Of ', round(max_steps$avg_steps, 1), ' Steps \n On ', max_steps$interval, 'th Time Interval', sep = '')

points(max_steps$interval,  max_steps$avg_steps, col = 'red', lwd = 3, pch = 19)
legend("topright",
       legend = max_lab,
       text.col = 'red',
       bty = 'n'
       )

```

---

## Missing Values

Number of Missing Values

```{r}
sum(is.na(dat$steps))
```
Fitting of missing values in dataset
Creation of new dataset

```{r}

setkey(dat_tbl, interval)
setkey(dat_tbl_summary_intv, interval)
NA_replace = function(x,y){
        if(is.na(x)){
                
                return(y)
        }
        return(x)
}

dat_tbl_miss = dat_tbl[dat_tbl_summary_intv]
dat_tbl_miss$new_steps = mapply(NA_replace,dat_tbl_miss$steps, dat_tbl_miss$avg_steps)

dat_tbl_summary_miss = dat_tbl_miss[, list(new_steps = sum(new_steps, na.rm = T)), 
                          by = date]

head(dat_tbl_summary_miss)
```

Making of histogram for total number of steps taken per day, and the calculation of mean & median. 


```{r}

gen_hist(dat_tbl_summary$total_steps, 'Missing Values Removed')
gen_hist(dat_tbl_summary_miss$new_steps, 'Missing Values Replaced With \n Mean For Interval')

```

**Inference** : After replacement of missing values, the median increased.

## Checking differences in activity patterns 
---
Creation of new dataset variable

```{r}
#Make Function To Return Either "Weekday" or "Weekend"
weekpart = function(x){
        if(x %in% c('Saturday', 'Sunday')){
                return('Weekend')
        }
        
        return('Weekday')
}

dat_tbl_miss$dayname = weekdays(as.Date(dat_tbl_miss$date))

dat_tbl_miss$daytype = as.factor(apply(as.matrix(dat_tbl_miss$dayname), 1, weekpart))

dat_tbl_summary_miss = dat_tbl_miss[, list(avg_steps = mean(new_steps, na.rm = T)), 
                          by = list(interval, daytype)]

#inspect dataset
str(dat_tbl_summary_miss)
```
```{r}
library(lattice)
xyplot(avg_steps~interval | daytype, data = dat_tbl_summary_miss,
      type = 'l',
      xlab = 'Interval',
      ylab = 'Number of Steps',
      layout = c(1,2))
```

